Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis \thanks{This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.}

### 2018-04-05 jhir/latex-manuscript/manuscript.tex
update manuscript

\author{Mehedi Hasan, PhDc$^1$\and 
April Idalski Carcone, PhD$^2$\and 
Sylvie Naar, PhD$^3$\and 
Susan Eggly, PhD$^4$\and  
Gwen L. Alexander, PhD$^5$\and 
Kathryn E Brogan Hartlieb, PhD$^6$\and 
Alexander Kotov, PhD$^1$
}

@@

\begin{abstract}
Motivational Interviewing (MI) is a communication technique to increase intrinsic motivation for behavior change. Although there is a strong empirical evidence linking “MI-consistent” counselor behaviors and patient statements of intrinsic motivation (i.e., “change talk”) for behavior change, the specific patterns of patient- counselor communication that are effective for eliciting patient change talk are a subject of ongoing research in behavioral sciences. A significant barrier to this research is the time and resource-intensive cognitive tasks, such as manual coding and sequential analysis of coded motivational interview transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had previously been annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify meaningful and interpretable counselor communication strategies to guide clinical practice.
\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
%\PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change [129]. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change which is accomplished through the exercise of relational and technical skills. [2]CITE] The relational hypothesis suggests that counselors’ use accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change. MI’s technical hypothesis [3]cite] states counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”. Cto increase intrinsic motivation (engaging in an activity for reasons of satisfaction rather than external consequences) atients“change talk”, that is, is the  statements patients make during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Change talk expressed during treatment sessions consistently predicts behavior change [440] with results persisting as long as 34 months post-intervention [541]. MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) will lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk. [3] These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal sequencing of patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

[6][3]Sequential analysis is an analytic approach to examine the temporal sequencing of behavioral events. [7,8].ADD] In a study of adults seeking or mandated to treatment for alcohol abuse, Moyers and Martin [9]ADD] found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. In a follow up study with the same population, [10]ADD] change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk. MIIN was unrelated to sustain talk but decreased the likelihood of change talk. Among young adults in brief motivational interviewing for hazardous alcohol consumption, Gaume and colleagues [11]ADD] also found a MICO-change talk and MICO-sustain talk sequential pattern; a MIIN-sustain talk pattern was not observed. A second study with the same population confirmed MICO led to significantly more change talk and sustain talk [12]ADD]. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior that led to increased change talk; reflections and other MICO behaviors but not questions were related to increased sustain talk. Glynn and colleagues [13]ADD] found reflections of change talk were more likely to elicit change talk among incarcerated adolescents with high rates of alcohol and marijuana use. Similarly, reflections of sustain talk were more likely to elicit more sustain talk. In contrast, Carcone and colleagues [14]ADD] found open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy were the only counselor behaviors likely to result in change talk among adolescents engaged in weight loss treatment. In a parallel study of the adolescents’ caregivers, Jacques-Tiura et al. [15]ADD] found the same; questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors’ use of reflections was the only MICO behavior consistently linked to change talk suggesting a need for additional research to understand in what contexts the various MICO strategies are effective, i.e., lead to change talk.

The sequential analysis procedure used in the above MI process studies [16-19] is based on first-order Markov Chain models [9, 10, 12]. Markov Chain is a discrete-time stochastic process with the property that the system state or condition changes over time and only depends upon previous event. Markov chain models have two main drawbacks. The first is their inability to preserve the long-term dependencies between observations in a sequence (they consider each observation to be dependent only on the immediately preceding one). In MI and other behavioral interventions, the antecedents of a given observed behavior could be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes, although in MI multiple similar behaviors may have the same outcome. Thus, first-order Markov models may be insufficient for understanding the associations among behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of behavior codes and long-term dependencies among behaviors, to identify such causal relationships. The goal of the current research is to develop a such a method to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting topic models [20-24], classification models [25-28] and neural networks [25, 29, 30] to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al. [27] developed a system for the evaluation of counselor fidelity to the MI framework by observing counselor language during the motivational interview. This system employed a support vector machine (SVM), which constructs a hyperplane in a high dimensional space to separate data points. In their approach, Perez-Rosas et al. proposed SVM model based on n-gram, syntactic, and semantic features, which achieved near 90\% accuracy for predicting specific MI counselor behaviors, such as reflections. Patient’s engagement, verbal and non-verbal accommodation as well as topics discussed during the MI session, were analyzed to identify linguistic and acoustic features of counselor empathy. These features were then used to build a counselor empathy classifier that achieved an accuracy of up to 80\%. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and neural networks in conjunction with lexical, contextual, and semantic features for the task of automated annotation of MI transcripts with different numbers of behavior codes (codebooks) [25].  In a follow up study [31], we formulated the problem of predicting the likelihood of eliciting a certain type of behavioral response during motivational interview as a sequence classification problem. Markov chains, HMMs and recurrent neural networks were applied to monitor the progression of motivational interviews and predicting the likelihood of eliciting “change talk”. This paper continues this line of research by examining the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the antecedent counselor communication strategies leading to patient change talk.

HMMs are widely used for analysis of sequence data due of their ability to model dependencies between coherent clusters of discrete observations in a sequence. HMM associates a “hidden” state with each observation in a sequence, where each state corresponds to a different distribution over observations, and model’s sequences of observations as transitions between hidden states and sampling observations from each hidden state. HMMs were originally proposed for speech recognition [3263], in which the states were used to represent all sounds of English. In biomedical informatics, HMMs were employed for diagnosis of diseases and biological sequence modeling [3366, 3467]. For example, in [3366] the Doppler ultrasound was used for feature extraction and then an HMM based classifier was applied to distinguish healthy patients from the patients with a heart disease, while in [3467], HMM was applied to capture important statistical characteristics of protein families. HMM and adjusted regression models were employed in [35] to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention.  In the case of applying HMM to modeling patient-counselor communication, hidden states correspond to a set of related behavior codes, such as a patient’s underlying emotional state, during a motivational interview. Although MI literature clearly identifies patient change talk and commitment language as antecedents of patient’s behavior change [40], there is less clarity on which counselor communication strategies influence the articulation of change talk. HMM modeling of successful and unsuccessful MI sessions could provide additional evidence to identify counselor communication strategies that contribute to patient change talk.

Frequent Pattern Mining [3668] is a class of data mining methods to identify sets of items (referred to as itemsets) which frequently appear together in a collection of sequences. Frequent Pattern Mining was first introduced by Agrawal and Srikant with the Apriori algorithm [3769], which was designed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied in several other domains including health informatics [3870-4173], medical imaging [39] and chemical and biological analysis [4274-4476], web mining [4577], and outlier analysis [4678]. Abdullah et. al. applied frequent pattern mining on medical billing data to analyze associations between diagnosis and treatments [3870]. In particular, the Apriori algorithm was used to identify most frequent itemsets1 that consist of a minimum of one diagnosis code (DxCode) and one current procedural terminology (CPT) code. Another study [3971] was conducted on medical image data for mining association rules to provide diagnosis support. At first, important features were extracted and then domain knowledge was integrated to generate rules for the interpretation of frequent patterns among important image features, such as object size, noise level, contrast, and texture, etc. However, to the best of our knowledge, nNo publishedrior research has yet examined the utility of frequent pattern mining to study patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which are difficult to interpret, is a major challenge in applying frequent pattern mining methods to patient- communication sequences. To address this problem, in this study, we utilized the closed frequent itemset mining method [4779], which produces more compact patterns that are easier to interpret. Specifically, we leveraged FPClose [4880], an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that most frequently lead to patient change talk. FPClose is the best algorithm in terms of running time and memory consumption proposed to date for mining closed frequent itemset, which works particularly well for sparse data.

This paper is one of the first empirical evaluations of the utility of HMM and closed frequent pattern mining for the study of patient-counselor communication during informing MI practice via automated analysis of patient- communication. Bertholet and colleagues published the only other study using this approach…. The goal of this study is to use HMM and FPM to inform clinical practitioners of the novel type of analytical results that these models can provide to help them better understand the antecedent counselor communication strategies leading to patient change talk during Motivational Interviewing sessions. In particular, tThese two models offerprovide the following advances over the sequential analysis computational methods currently used for MI research:. 
As opposed to fFirst-order Markov chain models, which identify the influence of counselor statements immediately preceding on patient behavior only in the form of statements using a transition matrix between of individual behavior codes. In contrast, HMMs identifies summarizeidentify transitions between clusters of related behavior codes, allowing theo capture identification to examine which of the antecedent behavior cluster is responsible for the elicitation of change talk and sustain talk in successful and unsuccessful patient-counselor communication.of more abstract patterns in patient-counselor communication;. 
As opposed to the first-order Markov models and HMMs, fFrequent pattern mining can capture the can identify patterns that involveing long-range dependencies between patient and counselor behaviors. Long-range dependency is important because human behavior is informed by all the antecedent behaviors not just the immediately preceding behavior.

In this paper, we focus on computational methods to facilitate analysis of transcripts to identify patterns of patient-counselor communication, in the form of sequences of behavior codes, in both successful and unsuccessful motivational interviews. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge can inform clinical practice by facilitating the development of more effective and tailored behavioral interventions, including the ones focused on weight loss among minority adolescents.
\section{Section title}
\label{sec:1}
Text with citations \citep{wei2006semi} and \citep{keogh2000scaling}.
\subsection{Subsection title}
\label{sec:2}
as required. Don't forget to give each section
and subsection a unique label (see Sect.~\ref{sec:1}).
\paragraph{Paragraph headings} Use paragraph headings as needed.
\begin{equation}
a^2+b^2=c^2
\end{equation}

% For one-column wide figures use
\begin{figure}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics{example.eps}
% figure caption is below the figure
\caption{Please write your figure caption here}
\label{fig:1}       % Give a unique label
\end{figure}
%
% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=0.75\textwidth]{example.eps}
% figure caption is below the figure
\caption{Please write your figure caption here}
\label{fig:2}       % Give a unique label
\end{figure*}
%
%\newgeometry{margin=2cm} % modify this if you need even more space
\begin{landscape}

% For tables use
\begin{table}
% table caption is above the table
\caption{Please write your table caption here}
\label{tab:1}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

% For tables use
\begin{table}
% table caption is above the table
\caption{Please write your table caption here}
\label{tab:2}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

>>>>

\author{Mehedi Hasan, BS$^1$\and 
April Idalski Carcone, PhD$^2$\and 
Sylvie Naar, PhD$^3$\and 
Susan Eggly, PhD$^4$\and  
Gwen L. Alexander, PhD$^5$\and 
Kathryn E Brogan Hartlieb, PhD$^6$\and 
Alexander Kotov, PhD$^1$
}

@@

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking “MI-consistent” counselor behaviors and patient motivational statements (i.e., “change talk”), the specific counselor communication behaviors effective for eliciting patient change talk are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining, an algorithm used to mine the closed frequent patterns in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had previously been annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify meaningful and interpretable counselor communication strategies to guide clinical practice.

\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
%\PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \citep{miller2013motivational}. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \citep{miller2013motivational}. The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change. MI’s technical hypothesis \citep{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”. Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4]. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5]. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations [6, 7]. Moyers and Martin [8] applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk [9]. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al. [10] used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk [11]. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues [12] linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al. [13] used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy. A parallel study of the adolescents’ caregivers [14] drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors’ use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.

The sequential analysis procedure used in the above MI process studies [8, 15-17] is based on the first-order Markov Chain model [8, 9, 11]. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of behavior codes and long-term dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of data mining methods to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and primarily for assessing intervention fidelity. Perez-Rosas et al. [25] developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, which achieved nearly 90\% accuracy for predicting counselor questions and reflections. Patient’s engagement, verbal and non-verbal accommodation as well as topics discussed during the MI session, were analyzed to identify linguistic and acoustic features of counselor empathy. These features were then used to build a counselor empathy classifier that achieved an accuracy of up to 80\%. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes [23]. The SVM model with the proposed features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders. In a follow up study [29], we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. This paper represents the next phase in this line of research. We examine the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the counselor communication strategies antecedent to patient change talk.

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a “hidden” state, where the hidden state corresponds to a different distribution over observations (i.e., distribution over behavior codes). Sequences of observations are modeled as transitions between different hidden states and sampling observations from each hidden state. HMMs were originally proposed for speech recognition [30], in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling [31, 32]. For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease [31], while HMM was used to capture important characteristics of protein families [32]. In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change [33]) as the antecedents of patient’s behavior change [3], there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.

Frequent pattern mining [34] is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant [35] first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics [36-39], medical imaging [37], chemical and biological analysis [40-42], web mining [43], and outlier analysis [44]. No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method [45], which produces less number of patterns in more compact form that are easier to interpret. In this study, we leveraged FPClose [46], an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. FPClose is the best closed frequent itemset mining algorithm, which shown highest performance in terms of running time and memory consumption.

In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI’s technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al. [47] used HMM and adjusted regression models to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention. Their HMM model limited hidden states to three, which are "towards change", "away from change" and "non-determined". First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communication during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

\section{Methods}
\label{sec:methods}
\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss treatment. Each transcript represented a one-on-one interaction between the adolescent patient and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE) [13]. The MYSCOPE is an adaptation of the original MI-SCOPE [48], a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI) [49], the MI Skill Code (MISC) [50], and Amrhein’s conceptualization of change talk and commitment [51]. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% (n = 7) of the sessions for inter-rater reliability which was good (k = .696). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

% For tables use
\begin{table}
% table caption is above the table
\caption{MYSCOPE codebook.}
\label{tab:codebook}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from beginning to end was analyzed. Successful sequences were those that resulted in a patient change talk or commitment language statement and all preceding behavior codes were included. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.

\subsection{Data Analysis}
\label{subsec:dataanalysis}
The Hidden Markov Model (HMM; implemented in the hmmlearn package which is publicly available at http://hmmlearn.readthedocs.io/en/latest/index.html) analysis aimed to identify clusters of related behavior codes within successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters. Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence. The emission probability matrix describes the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions among the hidden states.

Training an HMM with a given number of hidden states (N) involves estimating the following parameters using the Baum-Welch algorithm:
M is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table 1).
T is an  state transition probability matrix where  
E is an  matrix where , with observation symbol probability distribution for each state (emission probability matrix), and
p is the initial state distribution vector where .

In this study, two HMM models were trained, one using successful sequences and the other using unsuccessful sequences, with the objective of maximizing log-likelihood on the observed sequence. The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest BIC value, which indicates an optimal fit of HMM model. BIC minimizes negative log-likelihood but penalizes complexity (i.e. \# of hidden states), which is important for avoiding overfitting problem. HMM analyses suggested 5 hidden states was optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).

% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
% figure caption is below the figure
\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states.}
\label{fig:bic}       % Give a unique label
\end{figure*}

A second experiment was conducted with Frequent Pattern Mining. Orange3 and arules are two available pattern mining libraries implemented in python and r-programming, which support only a few pattern mining algorithms including frequent itemsets and association rules mining. However, we conducted our experiment using SPMF [52, 53], which offered more than 150 data mining algorithms. Moreover, SPMF is an open-source, fast and lightweight, a most popular and widely used Java-based data mining library. A frequent itemset is determined by surpassing a user-specified threshold called minimum support count\footnote{Support count: \# of times an itemset appears in the database}. For example, \{A\}, \{C\}, {D}, and {C, D} are frequent itemset in the database illustrated in Figure~\ref{fig:fpm}, because these itemsets appear at least 2 times which was the minimum support count identified in the example database. We reported closed frequent itemsets among all sequences of behavior codes in successful and unsuccessful sequences. A frequent itemset is closed if none of its supersets have the same support count [45], where a set X is a superset of another set Y if X contains all elements of the set Y. For example, the itemsets {A}, {C}, {D}, and {C, D} are not closed frequent itemsets because their supersets \{A, B\}, {B, C}, {B, D}, and {B, C, D} have the same number of support. Therefore, {B}, {A, B}, {A, D}, {B, C}, {B, D}, and {B, C, D} are closed frequent itemsets because none of their supersets have same the number of support count. On the other hand, {A, C}, {A, B, C}, {A, B, D}, {A, C, D} and {A, B, C, D} have support count 1, 0, 1, 0, and 1, respectively, which are less than minimum support count and identified as non-frequent itemsets. This study applied FPClose [46], an efficient close frequent pattern mining algorithm in SPMF, to identify frequent patient-counselor behaviors in MI communication sequences. To improve the model’s ability to identify frequent patient behaviors, FPClose [46], a FP-tree based algorithm [54, 55] was used. FPClose identifies closed frequent itemsets by utilizing an FP-tree data structure in combination with the FP-array technique to avoid the frequent traversal of FP-trees. Before applying the FPClose method, the dataset was transformed into the expected format used by the algorithm in the SPMF library. Minimum support threshold is a certain proportion having a meaningful representation of patterns determined by the domain expert. Since there is no standard for the selection of minimum support threshold, by following the studies [56-57], the minimum support count considered was 10\% of the total number of sequences which is 110 for successful and 25 for unsuccessful sequences. In our study, support counts representing ≥10\% of the patterns was the threshold for the result interpretation.

% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=0.95\textwidth]{figures/fpm.eps}
% figure caption is below the figure
\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
\label{fig:fpm}       % Give a unique label
\end{figure*}

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model, are reported in Tables~\ref{tab:transition} and~\ref{tab:emission}. The three behaviors with the highest frequency, representing 45-69\% of each state’s emission transitions, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began in a state characterized as “high motivation” as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began in a state of “high receptivity” 11\% of the time. “High receptivity” sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of “moderate receptivity” and “low receptivity” (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. Moderate receptivity sequences were distinguished from low receptivity sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the “active feedback” state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from “high motivation” to “active feedback” most often (41\%). “active feedback”, in turn, most frequently transitioned to “high receptivity” (39\%). The “moderate receptivity” state transitioned to most often to “active feedback” (27\%) and automatically transitioned back to “moderate receptivity” (24\%) or to “low receptivity” (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

In contrast, the majority of unsuccessful sequences (98\%) began in a state of “ambivalence” as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of “avoidance”. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished “avoidant” sequences. Both “ambivalent” (84\%) and “avoidant” (61\%) states most frequently transitioned to the “avoidant” state.

%\newgeometry{margin=2cm} % modify this if you need even more space
\begin{landscape}
% For tables use
\begin{table}
% table caption is above the table
\caption{Hidden Markov Model Emission Matrices}
\label{tab:emission}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

% For tables use
\begin{table}
% table caption is above the table
\caption{Hidden Markov Model Transition Matrices}
\label{tab:transition}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}
